Why does female meiosis allow the opportunity for exploitation by selfish sperm? BETTER TITLE?

### 2013-09-20 writeup_and_notes/male_dependent_drive.tex
Yaniv comments through discussion on friday sept 20

Here, we develop population genetic models exploring the evolution of sperm influence on female meiotic drive. 
%In this article we explore through simple population genetic models the consequences of alleles that influence the outcome of female meiosis.
These models show that sperm modifiers of female meiotic drive are
unlikely to create a sustained conflict, making it unlikely that
female meiosis will evolve to resist them.
In fact, the interests of sperm and egg  genomes' are often aligned, as are both invested in the fate of the resultant zygote (as was speculated for the In locus \citep{Pomiankowski1993}).
Therefore, there is little selective benefit to females in preventing sperm to influence female meioses,
	indeed features of meiosis may evolve that facilitate the interaction of sperm with female meiosis.

@@

%YB VERSION
When a deleterious recessive green-bearded driver does spread, 
	it spends much time at low frequency, because when rare it is unlikely to meet a complimentary sperm. 
However, once common, a green-bearded driver spreads rapidly to fixation due to its
	positive frequency dependent behavior (Inset in Figure \ref{Invasion_space}).  
Unlike standard meiotic drive systems, a homozygous cost, cannot sustain polymorphism of a green-bearded driver 
	because it already overcame this cost just to enter the population.
The lack of a balanced polymorphism precludes the evolution of an allele that suppresses this form of meiotic drive.

@@

%YB VERSION
Inclusion of a heterozygous fitness cost (e.g. additive fitness cost) further constrains the evolution of a sperm-dependent driver. In contrast to a  standard female  driver which can invade from low frequency so long as drive in heterozygotes outweighs the fitness cost they suffer (XXX); when rare our sperm dependent driver pays the heterozygous
        fitness cost but only weakly drives because the right sperm
        are rare. 
This frequency independent fitness cost, and positive frequency dependent drive generates a bistable system. 
%Therefore, even very weak heterozygous fitness costs
%	prevent the spread of a very rare green-bearded driver, unless its drive is very strong. 
%With an additive cost, the dynamics of a green-bearded driver resembles that of an underdominant mutation (in that the system is bistable). 
Below a cutoff frequency the green-bearded driver deterministically decreases in frequency 
	because it pays the  heterozygous fitness cost  but drives ineffectively (see FigureS\ref{bistable_additive}). 
Above this frequency, the allele increases deterministically because the commonness of sperm carrying this allele 
	make drive effective enough to overcome the associated fitness cost.
This bistability prevents sperm-dependent drivers from invading 	
	reasonably sized populations, and assures that if they do invade they will rapidly fix.

While an allele specifically active in sperm could plausibly influence the outcome of female meiosis, the limited levels of gene expression in sperm \citep[e.g.][]{Joseph2004}
	suggest it may also be plausible to consider 
sperm may influence female meiosis via expression of the fertilizing male's
diploid genotype (perhaps due to proteins and RNAs packaged into the sperm).
This complicates the dynamics of the system, as the dominance of
  the genotype of the male effect on female drive must be considered as a parameter. 
While under some circumstances the dominance of the drive effect in
males can allow the 
maintainance of the allele a polymorphism, the parameter range that
permits polymorphism is very narrow (and quite sensitive to the
dominance of the male effect, see Figure S\ref{Effect_of_dominance}).

@@

Above, we explored a model were a driver that drove in females when signaled by a complementary allele in sperm.  
This complicated pleiotropic effect, while plausible is somewhat unlikely. 
Here, we examine an alternative scenario (and more realistic) in which
initially a driver, with no sperm-depedency, is at drive-viability equilibrium (with two alleles
$A$ and $B$ are the ancestral non-driver and driver alleles).
Subsequently a sperm-acting modifier of female meotic drive arises tightly
        linked to this balanced locus (effectively creating a third
        allele/haplotype at this locus).
Assuming tight linkage between female driver and sperm modifier is consistent with the nature of well characterized meiotic drive systems 
	which are often maintained as polymorphic inversions with numerous linked modifiers (CITE).

@@

We will first consider the case where a sperm-acting facilitator of
female meiotic drive arises very tightly linked to the $B$ background 
creating the third allele $B^{+}$ (i.e. the non-recombining haplotype of a
        driver allele and sperm-acting drive modifier). 
This new allele which acts in sperm to increase the effectiveness of both
the $B$ and  $B^{+}$ drive in $AB$ and $AB^{+}$ heterozygotes, see Figure \ref{Eggsperm_3_allele_cartoon} for a simple schematic of this drive model.  
Naively this $B^{+}$ allele might be able to increase in frequency in the
population as then it could be allowed to spread due to the additional
drive it causes.
That turns out to not be the case, 
because the sperm-acting drive promoter (the novel $B^{+}$ haplotype) 
is introduced when the ancestral driver ($B$) is at drive-selection balance, 
	$B'$ will immediately suffer a homozygous fitness cost.  
Worse yet, a novel $B^{+}$ haplotype most often helps 
the $B$  allele drive ($B'$ sperm meeting $AB$ eggs), simply because $B$ is initially more common than $B^{+}$.
As $B^{+}$ facilitates $B$ these interactions will frequently form 
	$BB^{+}$ heterozygote that suffers from the full
	effect fitness costs as $BB$.

Sperm-acting self promoting alleles are therefore at a profound disadvantage
in this scenario, even more so than under the previous two allele model.
We have found no parameter range of this
three allele system that allows the selfish sperm-acting allele $B^{+}$ to
invade the population. \yb{MATHS?}

While  sperm enhancement of a female drive cannot displace a polymorphic female driver, sperm based drive suppression - that is sperm-acting alleles that act to restore 
fairness of female meiosis can. 
 If these sperm-acting drive suppressors arise on
	the $A$ background (creating a third allele $A^{-}$) 
or are unlinked to the drive locus they readily invade a population segregating
for the drive system ($AB$, see Figure). The invasion of such allele
will lower the frequency of the original drive system (perhaps to zero)
and will spread to fixation if they do not carry strong fitness costs
(Figure \ref{Eggsperm_3_allele_cartoon} and \ref{Trajectories_of_supressors}A). \\

Intriguingly these sperm disruptors of female meiotic drive can spread
	even when they arise tightly linked to the original drive system (B), forming
	an novel third allele $B^{-}$ as above (Figures \ref{Eggsperm_3_allele_cartoon} and \ref{Trajectories_of_supressors}B). 
$B^{-}$ displaces the original drive system ($B$) because it drives when not costly, 
	but by disrupting drive when transmitted to sperm it avoids low-fitness drive homozygotes ($BB^-$ and $B^-B^-$. 
Surprising despite displacing the ancestral driver ($B$), self avoiding driver, $B^*$ it's equilibrium frequency is not always higher than that of $B$.

>>>>

Here, we develop population genetic models exploring the evolution of sperm influence on female meiotic drive. 
%In this article we explore through simple population genetic models the consequences of alleles that influence the outcome of female meiosis.
These models show that sperm modifiers of female meiotic drive are
unlikely to create a sustained conflict, and therefore 
female meiosis will have little incentive to evolve resistance to them.
In fact, the interests of sperm and egg  genomes' are often aligned, as both are invested in the fate of the resultant zygote (as was speculated for the In locus \citep{Pomiankowski1993}).
Therefore, there is little selective benefit to females in preventing sperm to influence female meioses,
	indeed features of meiosis may evolve that facilitate the interaction of sperm with female meiosis.

@@

%YB VERSION
When a deleterious recessive green-bearded driver does spread, 
	it spends much time at low frequency, because when rare it is unlikely to meet a complimentary sperm. 
However, once common, a green-bearded driver spreads rapidly to fixation due to its
	positive frequency dependent behavior (Inset in Figure \ref{Invasion_space}).  
%Unlike standard meiotic drive systems, a homozygous cost cannot sustain polymorphism of a green-bearded driver 
%	because it already overcame this cost just to enter the population.
This lack of a balanced polymorphism precludes the evolution of an allele that suppresses this form of meiotic drive.

@@

%YB VERSION
Inclusion of a heterozygous fitness cost (e.g. additive fitness cost) further constrains the evolution of a sperm-dependent driver. 
While a  standard female  driver invades from low frequency when the transmission advantage outweighs the fitness cost (XXX),
	a rare sperm dependent driver pays a fitness cost, but drives only weakly because complementary sperm are uncommon. 
This frequency independent fitness cost, and positive frequency dependent drive generates a bistable system. 
%Therefore, even very weak heterozygous fitness costs
%	prevent the spread of a very rare green-bearded driver, unless its drive is very strong. 
%With an additive cost, the dynamics of a green-bearded driver resembles that of an underdominant mutation (in that the system is bistable). 
Below a cutoff frequency the green-bearded driver deterministically decreases in frequency 
	because it pays the  heterozygous fitness cost  but drives ineffectively (see FigureS\ref{bistable_additive}). 
Above this frequency, the allele increases deterministically because the commonness of complementary sperm 
	make drive effective enough to overcome the associated fitness cost.
This bistability prevents sperm-dependent drivers from invading 	
	reasonably sized populations, and assures that if they do invade they will rapidly fix.

While an the allelic identity of sperm could plausibly influence the outcome of female meiosis, 
	limited gene expression in sperm \citep[e.g.][]{Joseph2004}
	suggest a model where sperm influence female meiosis via expression of the fertilizing male's
	diploid genotype (perhaps due to proteins and RNAs packaged into the sperm).
This adds an additional parameter to the system -- the dominance of the drive enhancer in heterozygous males, 
	when this dominance is high, a narrow sliver of parameter space sustaining a polymorphism emerges (see Figure S\ref{Effect_of_dominance})
 	\yb{presumably because positive frequency dependence diminishes with the dominance of the male effect driver.}
%complicates the dynamics of the system, as the dominance of
 % the genotype of the male effect on female drive must be considered as a parameter. 
%Some circumstances the dominance of the drive effect in
%males can allow the  maintenance of the allele a polymorphism, the parameter range that
%permits polymorphism is very narrow (and quite sensitive to the
%dominance of the male effect, see Figure S\ref{Effect_of_dominance}).

@@

\yb{********* DO WE NEED TO DO THE MALE GENOTYPE THING HERE TOO? *********}

Above, we explored a model were a driver that drove in females when signaled by a complementary allele in sperm.  
This complicated pleiotropic effect, while plausible is somewhat unlikely. 
Here, we examine an alternative scenario (and more realistic) in which
initially a driver, with no sperm-dependency, is at drive-viability equilibrium (with two alleles
$A$ and $B$ are the ancestral non-driver and driver alleles).
Subsequently a sperm-acting modifier of female meiotic drive arises tightly
        linked to this balanced locus (effectively creating a third
        allele/haplotype at this locus).
Assuming tight linkage between female driver and sperm modifier is consistent with the nature of well characterized meiotic drive systems 
	which are often maintained as polymorphic inversions with numerous linked modifiers (CITE).

@@

We consider a sperm-acting facilitator of
	female meiotic drive tightly linked to the $B$ background, 
	generating a third allele $B^{+}$ (i.e. the non-recombining haplotype of a
        driver allele and sperm-acting drive modifier). 
This new allele acts in sperm to increase the effectiveness of both
	the $B$ and  $B^{+}$ drive in $AB$ and $AB^{+}$ heterozygotes, see Figure \ref{Eggsperm_3_allele_cartoon} 
	for a simple schematic of this drive model.  
Naively, the $B^{+}$ may spread by capitalizing on the additional drive it causes; hover,  
%Naively the $B^{+}$ allele might be able to increase in frequency in the
%	population as then it could be allowed to spread due to the additional
%	drive it causes. 
	this is not the case for a few simple reasons. 
First, the sperm-acting drive promoter (the novel $B^{+}$ haplotype) 
	arises when the ancestral driver ($B$) is at drive-selection balance, 
	and therefore immediately suffers a homozygous fitness cost.  
Worse yet, a novel $B^{+}$ haplotype most often helps 
	the $B$  allele drive ($B^+$ sperm meeting $AB$ eggs), because $B$ is initially more common than $B^{+}$. 
As $B^{+}$ facilitates $B$ these interactions frequently form 
	$BB^{+}$ heterozygotes that suffer the same fitness costs as $BB$. 
Therefore, sperm-acting self promoting alleles experience a profound disadvantage
	in this scenario, even more so than under the previous two allele model. 
We have found no parameter range of this
	three allele system that allows the selfish sperm-acting allele $B^{+}$ to
	invade the population. \yb{MATHS?}

While  sperm enhancement of a female drive cannot displace a polymorphic female driver, sperm based drive suppression - 
	that is sperm-acting alleles that act to restore 
	fairness of female meiosis can. 
If these sperm-acting drive suppressors arise on
	the $A$ background (creating a third allele $A^{-}$) ;
	or are unlinked to the drive locus they readily invade a population segregating
	for the drive system ($AB$, see Figure). 
This allele lowers the frequency of the original drive system (perhaps to zero),
	and spreads to fixation if it does not carry strong fitness costs
	(Figure \ref{Eggsperm_3_allele_cartoon} and \ref{Trajectories_of_supressors}A). \\
\yb{This result is unsurprising, as drive suppressors unlinked to drivers or in repulsion phase with them usually invade polymorphic drive systems (CITE).}

More surprisingly, sperm disruptors of female meiotic drive can spread
	even when tightly linked to the original drive system (B), forming
	an novel third allele $B^{-}$ as above 
	(Figures \ref{Eggsperm_3_allele_cartoon} and \ref{Trajectories_of_supressors}B). 
$B^{-}$ displaces the original driver ($B$) by driving when not costly, 
	but by disrupting drive when transmitted to sperm it avoids low-fitness drive homozygotes ($BB^-$ and $B^-B^-$). 
Despite displacing the ancestral driver ($B$), self avoiding driver, $B^*$ it's equilibrium frequency is not always higher than that of $B$.
